Optimizing TensorFlow's Memory Swapping by Prioritizing Earlier Tensors

### 2019-06-17 paper.tex
Revise paper

    Based on our experiments, the optimization reduces up to around 3\% of training duration on certain cases.
\end{abstract}

@@

\section{Related Works} \label{sec:relatedworks}

"Training deeper models by GPU memory optimization on TensorFlow" by Meng, Chen, et al. (2017) explores some memory problems in training deep learning models and proposes some solutions specifics to TensorFlow. Firstly, it addresses the OOM error in a training. Then, the paper proposes memory swapping as a solution to the OOM problem, and presents its own implementation of memory swapping in TensorFlow. Lastly, it gives some ideas on optimizing memory swapping in the context of training deep learning models, specifically on trainings with backpropagation. The paper explores some other topics mostly unrelated to this paper.

Meng, Chen, et al's paper is the main reference of this paper, as many topics explored in both papers are similar. However, both papers differ in various ways. As we previously mentioned, Meng, Chen, et al implemented \textit{their own} version of memory swapping in TensorFlow and the paper explained some of the optimizations. In contrast, we optimize the memory swapping \textit{already present} in TensorFlow, using one of the optimization methods in Meng, Chen, et al's implementation.

Based on our observations, Meng, Chen, et al's implementation of memory swapping in TensorFlow is significantly different than TensorFlow's own memory swapping. Thus, we only use the core idea of an optimization method in Meng, Chen, et al's implementation, and adapt it to TensorFlow's own memory swapping.

Before describing the optimization idea, we will briefly explain the deep learning training. In a training with backpropagation, there are two phases: feed forward and backpropagation. In the feed forward phase, the process goes from the input layer to the output layer, calculating outputs of each (hidden) layer's neurons. Afterwards, in the backpropagation phase, the process goes in reverse from the output layer to the input layer, calculating gradients of each (hidden) layer's neurons, used to adjust model parameters so that the overall error is minimized \cite{nielsen}.

@@

Based on those observations, the optimization idea is as follows. Earlier outputs (called "tensors" afterwards, since these outputs are represented as tensors in TensorFlow) occupy the memory space the longer until reused. Therefore, earlier tensors the best candidates to be memory swapped, because they let more time for the memory swapper to move them from the GPU to the CPU, and back. Since tensors are moved asynchronously, this can increase the overlapping between computations and swappings, thus reducing the overall training duration \cite{meng}.

@@

There are two main takeaways. The first is that swapped tensors are the latest tensors. The second is that even if tensors are swapped asynchronously, they are only swapped in when needed. For reasons we explained in Section~\ref{sec:relatedworks}, earlier tensors are hypothetically the better candidates for memory swapping, the main reason is that they let the memory swapper to asynchronously swap them in, along with other computations. This can potentially increase the asynchronicity of the program, reducing the overall training duration.

@@

After discussing the optimization method in Section~\ref{sec:relatedworks} and how TensorFlow's memory swapping works in Section~\ref{sec:memoryswapping}, in this section we will describe how can the optimization method be implemented in TensorFlow's memory swapping.

@@

As can be seen, the first half of the code is exactly the same as the code in \autoref{lst:StackPopOp}. The difference is, aside of that, we also asynchronously swap in "future tensors", swapped tensors that will be needed at a later time. As we previously mentioned in Section~\ref{sec:relatedworks} and Section~\ref{sec:memoryswapping}, this is to increase the asynchronicity of the program, ultimately reducing the overall training duration. That, is core of the optimization.

@@

We use the \textit{char-rnn} model as described in \cite{karpathy}, implemented in TensorFlow using \texttt{tf.nn.dynamic\_rnn} (mentioned in Section~\ref{sec:memoryswapping}) with \texttt{swap\_memory=True} to allow memory swapping. In short, char-rnn is an RNN model that takes a sequence of characters, then generates the next sequence of characters.

>>>>

    Based on our experiments on Char-RNN models with various hyperparameters and datasets (of size 285 KB and 4.4 MB), the optimization reduces up to around 3\% of training duration on certain cases.
\end{abstract}

@@

\section{Related Works}

\cite{meng} explores some memory problems in training deep learning models and proposes some solutions specifics to TensorFlow. Firstly, it addresses the OOM error in a training. Then, the paper proposes memory swapping as a solution to the OOM problem, and presents its own implementation of memory swapping in TensorFlow. Lastly, it gives some ideas on optimizing memory swapping in the context of training deep learning models, specifically on trainings with backpropagation.

This paper differs from \cite{meng} as the authors implemented \textit{their own} version of memory swapping in TensorFlow along with some of the optimizations. In contrast, we optimize the memory swapping \textit{already present} in TensorFlow, using one of the optimizations in \cite{meng}'s implementation, which, based on our observations, is significantly different than TensorFlow's own memory swapping. Thus, we only use the core idea of the optimization, and adapt it to TensorFlow's own memory swapping.

\section{Deep Learning Training and The Optimization} \label{sec:optimization}

Before describing the optimization idea, we will briefly explain about deep learning training. In a training with backpropagation, there are two phases: feed forward and backpropagation. In the feed forward phase, the process goes from the input layer to the output layer, calculating outputs of each (hidden) layer's neurons. Afterwards, in the backpropagation phase, the process goes in reverse from the output layer to the input layer, calculating gradients of each (hidden) layer's neurons, used to adjust model parameters so that the overall error is minimized \cite{nielsen}.

@@

Based on those observations, the optimization idea from \cite{meng} is as follows. Earlier outputs (called "tensors" afterwards, since these outputs are represented as tensors in TensorFlow) occupy the memory space the longer until reused. Therefore, earlier tensors are the best candidates to be memory swapped, because they let more time for the memory swapper to move them from the GPU to the CPU, and back. Since tensors are moved asynchronously, this can increase the overlapping between computations and swappings, thus reducing the overall training duration \cite{meng}.

@@

There are two main takeaways. The first is that swapped tensors are the latest tensors. The second is that even if tensors are swapped asynchronously, they are only swapped in when needed. For reasons we explained in Section~\ref{sec:optimization}, earlier tensors are hypothetically the better candidates for memory swapping, the main reason is that they let the memory swapper to asynchronously swap them in, along with other computations. This can potentially increase the asynchronicity of the program, reducing the overall training duration.

@@

After discussing the optimization method in Section~\ref{sec:optimization} and how TensorFlow's memory swapping works in Section~\ref{sec:memoryswapping}, in this section we will describe how can the optimization method be implemented in TensorFlow's memory swapping.

@@

As can be seen, the first half of the code is exactly the same as the code in \autoref{lst:StackPopOp}. The difference is, aside of that, we also asynchronously swap in "future tensors", swapped tensors that will be needed at a later time. As we previously mentioned in Section~\ref{sec:optimization} and Section~\ref{sec:memoryswapping}, this is to increase the asynchronicity of the program, ultimately reducing the overall training duration. That, is core of the optimization.

@@

We use the \textit{Char-RNN} model as described in \cite{karpathy}, implemented in TensorFlow using \texttt{tf.nn.dynamic\_rnn} (mentioned in Section~\ref{sec:memoryswapping}) with \texttt{swap\_memory=True} to allow memory swapping. In short, Char-RNN is an RNN model that takes a sequence of characters, then generates the next sequence of characters.
